Legacy Data Confounds Genomics Studies

### 2019-07-09 LatexManuscript/legacy.tex
removed some blue text, fixed S8

Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used to impute untyped variants, to estimate allele frequencies, and to answer a wide range of medical and evolutionary questions.
This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
\todo{Mafessoni et al. recently identified batch effects in the 1kGP by looking for individuals with excess LD among distant variants. \citep{mafessoni2018turning}.}
Here we point out how these and additional unreported batch effects in the early phases of the 1kGP lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.

@@

\todo{The mutational 3-mer substitution patterns of this list of variants from the GCAT model is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
There is also an enrichment of mutations from TAC, TCT, and TGT to the homonucleotide TTT. }\sgcomment{Point to figure}

@@

\begin{table}[h!]
\centering
\begin{tabular}{l  r r r  r r}
                      & {Repeat} & \textcolor{gray}{Proportion} & {Non-Repeat} & \textcolor{gray}{Proportion}   & {Total}    \\ \hline
{SNP}  & 3,369 &\textcolor{gray}{0.53\textperthousand} & 11,059 & \textcolor{gray}{0.56\textperthousand} & 14,428\\  
{INDEL} & 181& \textcolor{gray}{0.3\textperthousand} & 657& \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
{Total} & 3,550& & 11,716 & & 15,270\\
\end{tabular}
\caption{Number and proportion of $Q$-associated variants passing the 1000 Genomes Project strict mask per category. 
Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
A total of 15,270 are statistically significantly associated to $Q$. 
\todo{The grey text is the proportion of $Q$-associated variants per category.}
The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
\label{sigTable}
\end{table}

\subsection{Cell line or technical artifact}
\todo{
To assess whether $Q$-associated variants result from sequencing artifacts, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
Variants that do not reproduce across cohorts are likely the result of sequencing artifacts. 
Variants that do reproduce across studies could result from (recurrent) cell line mutations or, more likely, from existing variants whose accurate genotyping depends on $Q$.   
}

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=15cm,keepaspectratio]{../Figures/VariantDensity.jpg}
\caption{The number of $Q$-associated variants per 10kb window across the genome. \sgcomment{Any reason why this is plotted until 35?}
Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.}  
\label{VariantDensity}
\end{figure}

>>>>

Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used to impute untyped variants, to estimate allele frequencies, and to answer a wide range of medical and evolutionary questions.
This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
Mafessoni et al. recently identified batch effects in the 1kGP by looking for individuals with excess LD among distant variants. \citep{mafessoni2018turning}.
Here we point out how these and additional unreported batch effects in the early phases of the 1kGP lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.

@@

The mutational 3-mer substitution patterns of this list of variants from the GCAT model is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
There is also an enrichment of mutations from TAC, TCT, and TGT to the homonucleotide TTT (Figure \ref{strict_gcat_mutational_enrichment}).

@@

\begin{table}[h!]
\centering
\begin{tabular}{l  r r r  r r}
                      & {Repeat} & \textcolor{gray}{Proportion} & {Non-Repeat} & \textcolor{gray}{Proportion}   & {Total}    \\ \hline
{SNP}  & 3,369 &\textcolor{gray}{0.53\textperthousand} & 11,059 & \textcolor{gray}{0.56\textperthousand} & 14,428\\  
{INDEL} & 181& \textcolor{gray}{0.3\textperthousand} & 657& \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
{Total} & 3,550& & 11,716 & & 15,270\\
\end{tabular}
\caption{Number and proportion of $Q$-associated variants passing the 1000 Genomes Project strict mask per category. 
Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
A total of 15,270 are statistically significantly associated to $Q$. 
The grey text is the proportion of $Q$-associated variants per category.
The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
\label{sigTable}
\end{table}

\subsection{Cell line or technical artifact}
To assess whether $Q$-associated variants result from sequencing artifacts, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
Variants that do not reproduce across cohorts are likely the result of sequencing artifacts. 
Variants that do reproduce across studies could result from (recurrent) cell line mutations or, more likely, from existing variants whose accurate genotyping depends on $Q$.

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=15cm,keepaspectratio]{../Figures/VariantDensity.jpg}
\caption{The number of $Q$-associated variants per 10kb window across the genome.
Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.}  
\label{VariantDensity}
\end{figure}
